---
title: "Lecture 7<br/>Univariate analysis"
author: "Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/><a href=\"https://github.com/sdesabbata/GY7702\">github.com/sdesabbata/GY7702</a> licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    template: ../Utils/IOSlides/UoL_Template.html
    logo: ../Utils/IOSlides/uol_logo.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
rm(list = ls())
```



# Lecture 7<br/>Summary



## Previous lectures

TODO



## Today

TODO



# Lecture 7<br/>Correlation



## Correlation



## Pearson’s r



## Spearman’s coefficient



## Kendall’s tau



## Pairs plot



# Lecture 7<br/>Data transformations


# Z-scores



## Log transformation



## Inverse hyperbolic sine



# Lecture 7<br/> Regression



## Least squares



## Example

```{r, echo=TRUE}
library(tidyverse)
library(nycflights13)

nycflights13::flights %>%
  filter(month == 11, carrier == "US") %>%
  lm(dep_delay ~ arr_delay, data = .) %>%
  summary()

#filter(!is.na(dep_delay)) %>%
```



## Overall fit



## Parameters

Beta



## Outliers and residuals



## Influential cases



## Checking assumptions

<!--
Note from Discovering Statistics Using R by Andy Field, Jeremy Miles and Zoë Field, SAGE Publications Ltd, 2012
List all in 7.7.2.1 excluding those for multivariate
-->

## Normally distributed errors



## Homoscedasticity



## Independent errors



# Lecture 7<br/>Comparing means



## Independent t-test



## Reporting results



## Dependent t-test



## Reporting results



# Lecture 7<br/>Summary



## Summary

TODO



## Practical session

In the practical session we will see:

TODO