---
title: "Lecture 111<br/>Control and Functions"
author: "Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/><a href=\"https://github.com/sdesabbata/GY7702\">github.com/sdesabbata/GY7702</a> licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    template: ../Utils/IOSlides/UoL_Template.html
    logo: ../Utils/IOSlides/uol_logo.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
rm(list = ls())
```



# Recap @ 111



## Previous lectures

Moving from programming to data science

- Basic types and variables
- The pipe operator
- Coding style
- Complex data types
    - Vectors
    - Data Frames
- Data selection and filtering
- Join operations
- Table re-shaping



## This lecture

How to control the data processing flow

- Conditional statements
- Loops
    - While
    - For
- Functions
- Scope of a variable


# Conditional statements

## If
Format: **if** (*condition*) *statement*

- *condition*: expression returning a logic value (`TRUE` or `FALSE`)
- *statement*: any valid R statement
- *statement* only executed if *condition* is `TRUE`


```{r, echo=TRUE}
a_value <- -7
if (a_value < 0) cat("Negative")
a_value <- 8
if (a_value < 0) cat("Negative")
```


## Else
Format: **if** (*condition*) *statement1* **else** *statement2*

- *condition*: expression returning a logic value (`TRUE` or `FALSE`)
- *statement1* and *statement2*: any valid R statements
- *statement1*  executed if *condition* is `TRUE`
- *statement2*  executed if *condition* is `FALSE`


```{r, echo=TRUE}
a_value <- -7
if (a_value < 0) cat("Negative") else cat("Positive")
a_value <- 8
if (a_value < 0) cat("Negative") else cat("Positive")
```

<!--
## Example

```{r, echo=TRUE}
x <- 10
if (is.numeric(x) & (!(x < 0) & (x != 0)) ) cat("greater than zero")
```
-->


## Code blocks

Suppose you want to execute **several** statements within a function, or if a condition is true

- Such a group of statements are called **code blocks**
- `{` and `}` contain code blocks

```{r, echo=TRUE}
first_value <- 8
second_value <- 5
if (first_value > second_value) {
  cat("First is greater than second\n") 
  difference <- first_value - second_value
  cat("Their difference is ", difference)
}
```



# Loops

## Loops
Loops are a fundamental component of (procedural) programming.


There are two main types of loops:

- **conditional** loops are executed as long as a defined condition holds true
    - construct `while`
    - construct `repeat`
- **deterministic** loops are executed a pre-determined number of times
    - construct `for`


## While

The *while* construct can be defined using the `while` reserved word, followed by the conditional statement between simple brackets, and a code block. The instructions in the code block are re-executed as long as the result of the evaluation of the conditional statement is `TRUE`.

```{r, echo=TRUE}
current_value <- 0

while (current_value < 3) {
  cat("Current value is", current_value, "\n")
  current_value <- current_value + 1
}
```

<!--
## Repeat

The *repeat* construct can be defined using the `repeat` reserved word, followed by a code block. The instructions in the code block are re-executed until the command `break` is given. The latter is currently given through an `if` construct, which tests the condition that would stop the loop.

```{r, echo=TRUE}
current_value <- 0

repeat {
  cat("Current value is", current_value, "\n")
  current_value <- current_value + 1
  if (current_value >= 3) break
}
```


## While vs Repeat
The difference between `while` and `repeat` is mostly syntactical.

- Sometimes one or the other might fit better with the algorithm you have in mind
- Use the one that comes easier to you in the given situation
-->

## For

The *for* construct can be defined using the `for` reserved word, followed by the definition of an **iterator**. The iterator is a variable which is temporarily assigned with the current element of a vector, as the construct iterates through all elements of the list. This definition is followed by a code block, whose instructions are re-executed once for each element of the vector.

```{r, echo=TRUE}
cities <- c("Derby", "Leicester", "Lincoln", "Nottingham")
for (city in cities) {
  cat("Do you live in", city, "?\n")
}
```


## For

It is common practice to create a vector of integers on the spot in order to execute a certain sequence of steps a pre-defined number of times.

```{r, echo=TRUE}
for (i in 1:3) {
  cat("This is exectuion number", i, ":\n")
  cat("    See you later!\n")
}
```


## Loops with conditional statements

```{r, echo=TRUE}
3:0
#Example: countdown!
for (i in 3:0) {
  if (i == 0) {
    cat("Go!\n")
  } else {
    cat(i, "\n")
  }
}
```



# Functions


## Defining functions

A function can be defined 

- using an **identifier** (e.g., `add_one`) 
- on the left of an **assignment operator** `<-`
- followed by the corpus of the function

```{r, echo=TRUE}
add_one <- function (input_value) {
  output_value <- input_value + 1
  output_value
  }
```

## Defining functions

The corpus 

- starts with the reserved word `function`
- followed by the **parameter(s)** (e.g., `input_value`) between simple brackets
- and the instruction(s) to be executed in a code block
- the value of the last statement is returned as output

```{r, echo=TRUE}
add_one <- function (input_value) {
  output_value <- input_value + 1
  output_value
  }
```


## Defining functions

After being defined, a function can be invoked by specifying the **identifier**

```{r, echo=TRUE}
add_one (3)
```


## More parameters

- a function can be defined as having two or more **parameters** by specifying  more than one parameter name (separated by **commas**) in the function definition
- a function always take as input as many values as the number of parameters specified in the definition
    - otherwise an error is generated

```{r, echo=TRUE}
area_rectangle <- function (hight, width) {
  area <- hight * width
  area
}

area_rectangle(3, 2)
```


## Functions and control structures

Functions can contain both loops and conditional statements in their corpus

```{r, echo=TRUE}
factorial <- function (input_value) {
  result <- 1
  for (i in 1:input_value) {
    cat("current:", result, " | i:", i, "\n")
    result <- result * i
  }
  result
}
factorial(3)
```


<!--
## Functions and control structures

Factorial using `while`

```{r, echo=TRUE}
factorial <- function (input_value) {
  result <- 1
  while (input_value > 0) {
    cat("current:", result, " | input:", input_value, "\n")
    result <- result * input_value
    input_value <- input_value - 1
  }
  result
}
factorial(3)
```
-->



## Scope

The **scope of a variable** is the part of code in which the variable is ``visible''

In R, variables have a **hierarchical** scope:

- a variable defined in a script can be used referred to from within a definition of a function in the same scrip
- a variable defined within a definition of a function will **not** be referable from outside the definition
- scope does **not** apply to `if` or loop constructs


## Example

In the case below

- `x_value` is **global** to the function `times_x`
- `new_value` and `input_value` are **local** to the function `times_x`
    - referring to `new_value` or `input_value` from outside the definition of `times_x` would result in an error

```{r, echo=TRUE}
x_value <- 10
times_x <- function (input_value) {
  new_value <- input_value * x_value
  new_value
}
times_x(2)
```
<!--
Hint: try *debug* the function above and observe how the *local* environment is created and then discarded
-->


<!--
# Debugging

## What is debugging?

- Hardly any reasonably sized function works first time!
- Two broad kinds of problem
    - The function crashes (i.e. throws up an error)
    - The function doesn't crash  but returns the wrong answer
    - Generally, the second kind of error is the worst
- Prevention is better than cure, test the code as you write.
- **Debugging** is the process of finding the problems in the code.
- A typical approach
    - `Step' through the function line by line.  Find out where a crash occurs,  if one does.
    - Check the values of variables, and see if they are doing what they are supposed to.
- R has tools to help with this.


## To debug a function

- Enter `debug( <<Function Name>> )`
- For example: `debug(area.tri)`
- Then just use the function - it goes into 'debug mode'.
- Prompt becomes `Browse>`
- Line of function about to be executed is listed
- Pressing return executes it and goes to the next line
- Typing in a variable name lists the value of that variable
- Typing in any other command executes that command
- Typing in `c` and the return runs to the end of a function/block
- Typing in `Q` exits the function
- R can `see' variables that are specific to the function
- Enter `undebug(<<Function Name>>)` to return to normal


## Example

Debug the following example

```{r, echo=TRUE, eval=FALSE}
percentage_change <- function (before, after){
  difference <- before - after
  prop_change <- difference / before
  perc_change <- prop_change * 100
  perc_change
```
-->


# Summary



## Summary

How to control the data processing flow

- Conditional statements
- Loops
    - While
    - For
- Functions
- Scope of a variable
- Debugging



## Practical session

In the practical session, we will see

- Conditional statements
- Loops
    - While
    - For
- Functions
    - Loading functions from scripts
- Debugging



## Next lecture

- Data visualisation
    - histograms
    - boxplots
    - scatterplots
- Descriptive statistics
- Exploring assumptions
    - Shapiro–Wilk test
    - skewness and kurtosis
    - Levene’s test