---
title: "Lecture 203"
author: "Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/><a href=\"https://github.com/sdesabbata/GY7702\">github.com/sdesabbata/GY7702</a> licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    template: ../Utils/IOSlides/UoL_Template.html
    logo: ../Utils/IOSlides/uol_logo.png
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
rm(list = ls())
```



# Data manipulation



## Recap

**Prev**: Data selection and filtering

- dplyr
- dplyr::select
- dplyr::filter

**Now**: Data manipulation

- dplyr::arrange
- dplyr::summarise
- dplyr::group_by
- dplyr::mutate



## dplyr::arrange

```{r, echo=TRUE}
nov_dep_delays <- nov_dep_delays %>%
  arrange( 
    dest, # Ascending destination name
    -dep_delay # Descending delay
  )

nov_dep_delays[1:3, ]
```



## dplyr::summarise

`summarise`: calculate aggregated values (e.g., mean, max, etc)

```{r, echo=TRUE}
aggr_dep_delays_nov <- nov_dep_delays %>%
  # Need to filter out rows where delay is NA
  filter(!is.na(dep_delay)) %>%
  # Create two aggregated columns
  summarize(
    avg_dep_delay = mean(dep_delay), 
    tot_dep_delay = sum(dep_delay)
  )

aggr_dep_delays_nov
```



## dplyr::group_by

```{r, echo=TRUE}
dest_dep_delays_nov <- nov_dep_delays %>%
  # Need to filter out rows where delay is NA
  filter(!is.na(dep_delay)) %>%
  # Frist group by same destination
  group_by(dest) %>%
  # Then caluculate aggregated value
  summarize(
    tot_dep_delay = sum(dep_delay)
  )

dest_dep_delays_nov[1:3, ]
```



## dplyr::mutate

```{r, echo=TRUE}
dest_dep_delays_nov <- dest_dep_delays_nov %>%
  mutate(
    tot_dep_delay_days = ((tot_dep_delay / 60) /24)
  )

dest_dep_delays_nov[1:3, ]
```


## Full pipe example

```{r, echo=TRUE}
dest_dep_delays_nov <- nycflights13::flights %>%
  select(origin, dest, dep_delay, arr_delay, year:day) %>%
  select(-arr_delay) %>%
  filter(month == 11) %>%
  filter(!is.na(dep_delay)) %>%
  arrange(dest, -dep_delay) %>%
  group_by(dest) %>%
  summarize(tot_dep_delay = sum(dep_delay)) %>%
  mutate(tot_dep_delay_days = ((tot_dep_delay / 60) /24))
  
dest_dep_delays_nov[1:3, ]
```



## Summary

Data manipulation

- dplyr::arrange
- dplyr::summarise
- dplyr::group_by
- dplyr::mutate

**Next**: Practical session

- Creating R projects
- Creating R scripts
- Data wrangling script