---
title: "Lecture 322"
author: "Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/><a href=\"https://github.com/sdesabbata/GY7702\">github.com/sdesabbata/GY7702</a> licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    template: ../Utils/IOSlides/UoL_Template.html
    logo: ../Utils/IOSlides/uol_logo.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
rm(list = ls())
```

<style type="text/css">
.small_r_all pre{
  font-size: 16px;
  line-height: 18px;
}
.small_r_output pre:not(.prettyprint){
  font-size: 16px;
  line-height: 18px;
}
.verysmall_r_output pre:not(.prettyprint){
  font-size: 12px;
  line-height: 14px;
}
</style>


# Assessing regression assumptions


## Recap

**Prev**: Simple Regression

- Regression
- Ordinary Least Squares
- Fit

**Now**: Assessing regression assumptions

- Normality
- Homoscedasticity
- Independence


## Checking assumptions

- **Linearity**
    - the relationship is actually linear
- **Normality** of residuals
    - standard residuals are normally distributed with mean `0`
- **Homoscedasticity** of residuals
    - at each level of the predictor variable(s) the variance of the standard residuals should be the same (*homo-scedasticity*) rather than different (*hetero-scedasticity*) 
- **Independence** of residuals
    - adjacent standard residuals are not correlated
- When more than one predictor: **no multicollinearity**
    - if two or more predictor variables are used in the model, each pair of variables not correlated



## Normality

Shapiro-Wilk test for normality of standard residuals, 

- robust models: should be not significant 

<div class="columns-2">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
delay_model %>% 
  rstandard() %>% 
  shapiro.test()
```

<font size="4">	
**Standard residuals are NOT normally distributed**
</font>

<center>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 4}
delay_model %>% 
  rstandard() %>%
  data.frame(std_res = .) %>%
  ggplot(aes(x = std_res)) +
  geom_histogram(
    aes(
      y =..density..
    ),
    bins = 100
  ) + 
  stat_function(
    fun = dnorm, 
    args = list(
      mean = delay_model %>% rstandard() %>% mean(),
      sd = delay_model %>% rstandard() %>% sd()),
    colour = "red", size = 1)
```
</center>

</div>



## Homoscedasticity

Breusch-Pagan test for homoscedasticity of standard residuals

- robust models: should be not significant

<div class="columns-2 small_r_output">

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 3}
library(lmtest)

delay_model %>% 
  bptest()
```

<font size="4">	
**Standard residuals are homoscedastic**
</font>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 3}
delay_model %>% 
  plot(which = c(1))
```

</div>



## Independence

Durbin-Watson test for the independence of residuals

- robust models: statistic should be close to 2 (between 1 and 3) and not significant

<div class="small_r_output">

```{r, echo=TRUE}
# Also part of the library lmtest
delay_model %>%
  dwtest()
```

</div>

<font size="4">	
**Standard residuals might not be completely indipendent**

Note: the result depends on the order of the data.
</font>




## Summary

Assessing regression assumptions

- Normality
- Homoscedasticity
- Independence

**Next**: Assessing regression assumptions

- Normality
- Homoscedasticity
- Independence