---
title: "501 Exploratory analysis"
author: "Stefano De Sabbata"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
rm(list = ls())
```

# Exploratory analysis

*[Stefano De Sabbata](https://stefanodesabbata.com)*

[This work](https://github.com/sdesabbata/GY7702) is licensed under the [GNU General Public License v3.0](https://www.gnu.org/licenses/gpl-3.0.html). Contains public sector information licensed under the [Open Government Licence v3.0](http://www.nationalarchives.gov.uk/doc/open-government-licence).

## Introduction


```{r, echo=TRUE, eval=FALSE}
library(tidyverse)
library(knitr)
library(broom)
leicester_2011OAC <- read_csv("2011_OAC_Raw_uVariables_Leicester.csv")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(broom)
leicester_2011OAC <- read_csv("~/Repos/GY7702/Data/2011_OAC_Raw_uVariables_Leicester.csv")
```

`u011` Age 20 to 24

## Data visualisation

figures in RMarkdown `message=FALSE, warning=FALSE, fig.height = 4, fig.width = 4`

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.height = 3, fig.width = 6}
leicester_2011OAC %>%
  ggplot(
    aes(
      x = u011
    )
  ) +
  geom_histogram(binwidth = 5) +
  theme_bw()
```

`fig.height = 4` only

colour vs fill
https://ggplot2.tidyverse.org/reference/aes_colour_fill_alpha.html


```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.height = 3}
leicester_2011OAC %>%
  ggplot(
    aes(
      x = Total_Population,
      y = u011,
      colour = supgrpname
    )
  ) +
  geom_point(size = 1) +
  coord_fixed(ratio = 1) +
  theme_bw()
```

forcats, fct_reorder and 
colorbrewer
`\n`
  
```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.height = 3}
leicester_2011OAC %>%
  ggplot(
    aes(
      x = Total_Population,
      y = u011,
      colour = fct_reorder(supgrpname, supgrpcode)
    )
  ) +
  geom_point(size = 1) +
  scale_colour_brewer(palette = "Set1") +
  coord_fixed(ratio = 1) +
  guides(
    colour = guide_legend(title = "2011 Output Area Classification\n(supergroups)")
  ) +
  theme_bw()
```

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.height = 3, fig.width = 6}
leicester_2011OAC %>%
  mutate(
    perc_age_20_to_24 = (u011 / Total_Population) * 100
  ) %>%
  ggplot(
    aes(
      x = perc_age_20_to_24
    )
  ) +
  geom_histogram(
    binwidth = 1
  )+
  xlab("Percentage of residents aged 20 to 24") +
  theme_bw() 
```

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width = 6}
leicester_2011OAC %>%
  mutate(
    perc_age_20_to_24 = (u011 / Total_Population) * 100
  ) %>%
  ggplot(
    aes(
      x = supgrpname,
      y = perc_age_20_to_24
    )
  ) +
  geom_boxplot() +
  xlab("2011 Output Area Classification (supergroups)") +
  ylab("Percentage of residents aged 20 to 24") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.height = 5, fig.width = 6}
leicester_2011OAC %>%
  mutate(
    perc_age_20_to_24 = (u011 / Total_Population) * 100
  ) %>%
  ggplot(
    aes(
      x = fct_reorder(supgrpname, supgrpcode),
      y = perc_age_20_to_24,
      fill = fct_reorder(supgrpname, supgrpcode)
    )
  ) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set1") +
  xlab("2011 Output Area Classification (supergroups)") +
  ylab("Percentage of residents aged 20 to 24") +
  guides(
    fill = guide_legend(title = "2011 Output Area Classification\n(supergroups)")
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## Exercise 6.1

Create a RMarkdown analysis docuement that replicates the analysis above for another age group of your choice. Include conclusions that you can draw from each plot (about 100 words per plot). Also, `geom_bin2d()`

## Exploratory statistics

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(pastecs)
library(car)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 <- leicester_2011OAC %>%
  mutate(
    perc_age_20_to_24 = (u011 / Total_Population) * 100,
    supgrpname = dplyr::recode(supgrpname, 
      `Suburbanites` = "SU",
      `Cosmopolitans` = "CP",
      `Multicultural Metropolitans` = "MM",
      `Ethnicity Central` = "EC",
      `Constrained City Dwellers` = "CD",
      `Hard-Pressed Living` = "HP",
      `Urbanites` = "UR"
    )
  ) %>%
  select(OA11CD, supgrpname, perc_age_20_to_24)

leic_2011OAC_20to24 %>%
  top_n(5) %>%
  kable()
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 %>%
  select(perc_age_20_to_24) %>%
  stat.desc() %>%
  kable(digits = 3)
```

recode, dplyr vs car

```{r, echo=FALSE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24_supgrp_desc <- leic_2011OAC_20to24 %>%
  spread(
    key = supgrpname,
    value = perc_age_20_to_24
  ) %>%
  select(-OA11CD) %>%
  stat.desc(norm = TRUE)

leic_2011OAC_20to24_supgrp_desc %>%
  kable(digits = 3)
```

**Levene’s test** for equality of variance in different levels

- If significant, the variance is different in different levels

Used to test assumptions of regression, but also useful here.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leveneTest(leic_2011OAC_20to24$perc_age_20_to_24, leic_2011OAC_20to24$supgrpname)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 6, out.width='100%'}
leic_2011OAC_20to24 %>%
  filter(supgrpname == "CP") %>%
  ggplot(
    aes(
      x = perc_age_20_to_24
    )
  ) +
  geom_histogram(
    aes(
      y =..density..
    ),
    binwidth = 5
  ) + 
  stat_function(
    fun = dnorm, 
    args = list(
      mean = leic_2011OAC_20to24_supgrp_desc["mean", "CP"],
      sd = leic_2011OAC_20to24_supgrp_desc["std.dev", "CP"]),
    colour = "red", size = 1
  )
```