```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
rm(list = ls())
```



# Exploring assumptions


## Recap

**Prev**: Descriptive statistics

- stat.desc
- dplyr::across

**Next**: Exploring assumptions

- Normality
- Skewness and kurtosis
- Homogeneity of variance


## Libraries and data

```{r, echo=TRUE, message=FALSE, warning=FALSE,}
library(tidyverse)
library(magrittr)
library(knitr)

library(pastecs)

library(nycflights13)

flights_nov_20 <- nycflights13::flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay), month == 11, day ==20) 
```


## Normal distribution

<div class="columns-2">

- characterized by the bell-shaped curve 
- majority of values lie around the centre of the distribution
- the further the values are from the centre, the lower their frequency
- about 95% of values within 2 standard deviations from the mean

<br/><br/><br/>

```{r, echo=FALSE, fig.height=6, fig.width=6, out.width='100%'}
x <- seq(-4, 4, length=100)
hx <- dnorm(x)

plot(x, hx, type="l", xlab="Values",
  ylab="Density", main="Normal distribution")
```

</div>


## Density histogram

<div class="columns-2">

```{r, echo=FALSE, message=FALSE, warning=FALSE}
flights_desc_stats <- nycflights13::flights %>%
  filter(month == 11, carrier == "US") %>%
  select(dep_delay, arr_delay, distance) %>%
  stat.desc(norm = TRUE)

dep_delay_mean <- flights_desc_stats["mean", "dep_delay"]
dep_delay_sd <- flights_desc_stats["std.dev", "dep_delay"]
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 5.5, fig.width = 4, out.width='100%'}
nycflights13::flights %>% 
  ggplot(
    aes(
      x = dep_delay
    )
  ) +
  geom_histogram(
    aes(
      y =..density..
    ),
    binwidth = 10) + 
  stat_function(
    fun = dnorm, 
    args = list(
      mean = dep_delay_mean,
      sd = dep_delay_sd),
    colour = "red", size = 1)
```


```{r, echo=TRUE, eval=FALSE}
nycflights13::flights %>% 
  ggplot(
    aes(
      x = dep_delay
    )
  ) +
  geom_histogram(
    aes(
      y =..density..
    ),
    binwidth = 10
  ) + 
  stat_function(
    fun = dnorm, 
    args = list(
      mean = dep_delay_mean,
      sd = dep_delay_sd),
    colour = "black", size = 1)
```

</div>


## Q-Q plot

<div class="columns-2">

Cumulative values against the cumulative probability of a particular distribution

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 4}
nycflights13::flights %>%
  filter(
    month == 11, 
    carrier == "US"
  ) %>%
  qplot(
    sample = dep_delay, 
    data = .,
    stat = "qq", 
    xlab = "Theoretical", 
    ylab = "Sample"
  )
```

<br/><br/>

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights %>%
  filter(
    month == 11, 
    carrier == "US"
  ) %>%
  qplot(
    sample = dep_delay, 
    data = .,
    stat = "qq", 
    xlab = "Theoretical", 
    ylab = "Sample"
  )
```

</div>


## stat.desc: norm

```{r, echo=TRUE}
nycflights13::flights %>%
  filter(month == 11, carrier == "US") %>%
  select(dep_delay, arr_delay, distance) %>%
  stat.desc(basic = FALSE, desc = FALSE, norm = TRUE) %>%
  kable()
```



## Normality

**Shapiro–Wilk test** compares the distribution of a variable with a normal distribution having same mean and standard deviation

- If significant, the distribution is not normal
- `normtest.W` (test statistics) and `normtest.p` (significance)
- also, `shapiro.test` function is available

```{r, echo=TRUE}
nycflights13::flights %>%
  filter(month == 11, carrier == "US") %>%
  pull(dep_delay) %>%
  shapiro.test()
```



## Significance

Most statistical tests are based on the idea of hypothesis testing

- a **null hypothesis** is set
- the data are fit into a statistical model
- the model is assessed with a **test statistic**
- the **significance** is the probability of obtaining that test statistic value by chance

The threshold to accept or reject an hypotheis is arbitrary and based on conventions (e.g., *p < .01* or *p < .05*)

**Example:** The null hypotheis of the Shapiro–Wilk test is that the sample is normally distributed and *p < .01* indicates that the probability of that being true is very low.



## Skewness and kurtosis

In a normal distribution, the values of *skewness* and *kurtosis* should be zero

- `skewness`: **skewness** value indicates
    - positive: the distribution is skewed towards the left
    - negative: the distribution is skewed towards the right
- `kurtosis`: **kurtosis** value indicates
    - positive: heavy-tailed distribution
    - negative: flat distribution
- `skew.2SE` and `kurt.2SE`: skewness and kurtosis divided by 2 standard errors. If greater than 1, the respective statistics is significant (*p < .05*).



## Homogeneity of variance

**Levene’s test** for equality of variance in different levels

- If significant, the variance is different in different levels

```{r, echo=TRUE, message=FALSE, warning=FALSE}
dep_delay_carrier <- nycflights13::flights %>%
  filter(month == 11) %>%
  select(dep_delay, carrier)

library(car)
leveneTest(dep_delay_carrier$dep_delay, dep_delay_carrier$carrier)
```


## Summary

Exploring assumptions

- Normality
- Skewness and kurtosis
- Homogeneity of variance

**Next**: Practical session

- Data visualisation
- Descriptive statistics
- Exploring assumptions