```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
rm(list = ls())
```



<style type="text/css">
.small_r_all pre{
  font-size: 16px;
  line-height: 18px;
}
.small_r_output pre:not(.prettyprint){
  font-size: 16px;
  line-height: 18px;
}
.verysmall_r_output pre:not(.prettyprint){
  font-size: 12px;
  line-height: 14px;
}
</style>



# Correlation


## Recap

**Prev**: Comparing groups

- T-test
- ANOVA
- Chi-square

**Now**: Correlation

- Pearson’s r
- Spearman’s rho
- Kendall’s tau
- Pairs plot


## Correlation

Two variables can be related in three different ways

- related
    - positively: entities with high values in one tend to have high values in the other
    - negatively: entities with high values in one tend to have low values in the other
- not related at all

**Correlation** is a standardised measure of covariance



## Libraries and data

```{r, echo=TRUE, message=FALSE, warning=FALSE,}
library(tidyverse)
library(magrittr)  
library(nycflights13)

flights_nov_20 <- nycflights13::flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay), month == 11, day ==20) 
```



## Example

<center>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 4}
flights_nov_20 %>%
  ggplot(aes(x = dep_delay, y = arr_delay)) +
  geom_point() + coord_fixed(ratio = 1)
```
</center>


## Example

<div class="small_r_output">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
flights_nov_20 %>%
  pull(dep_delay) %>% shapiro.test()

flights_nov_20 %>%
  pull(arr_delay) %>% shapiro.test()
```

</div>



## Pearson’s r

<div class="columns-2 small_r_output">

If  two variables are **normally distributed**, use **Pearson's r**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
flights_nov_20_cor <- flights_nov_20 %$%
  cor.test(dep_delay, arr_delay)
```

The square of the correlation value indicates the percentage of shared variance

*If they were normally distributed, but they are not* 

- `r flights_nov_20_cor$estimate %>% round(digits = 3)` ^ 2 = `r flights_nov_20_cor$estimate ^ 2 %>% round(digits = 3)`
- departure and arrival delay *would* share `r ((flights_nov_20_cor$estimate ^ 2) * 100) %>% round(digits = 1)`% of variance

<br/><br/>

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# note the use of %$%
#instead of %>%
flights_nov_20 %$%
  cor.test(dep_delay, arr_delay)
```

</div>



## Spearman’s rho

<div class="columns-2 small_r_output">

```{r, echo=FALSE, message=FALSE, warning=FALSE}
flights_nov_20_cor_spearman <- flights_nov_20 %$%
  cor.test(dep_delay, arr_delay, method = "spearman")
```

If  two variables are **not normally distributed**, use **Spearman’s rho**

- non-parametric
- based on rank difference

The square of the correlation value indicates the percentage of shared variance

*If few ties, but there are*

- `r flights_nov_20_cor_spearman$estimate %>% round(digits = 3)` ^ 2 = `r flights_nov_20_cor_spearman$estimate ^ 2 %>% round(digits = 3)`
- departure and arrival delay *would* share `r ((flights_nov_20_cor_spearman$estimate ^ 2) * 100) %>% round(digits = 1)`% of variance


```{r, echo=TRUE, message=FALSE}
flights_nov_20 %$%
  cor.test(
    dep_delay, arr_delay, 
    method = "spearman")
```

</div>



## Kendall’s tau

<div class="columns-2 small_r_output">

```{r, echo=FALSE, message=FALSE, warning=FALSE}
flights_nov_20_cor_kendall <- flights_nov_20 %$%
  cor.test(dep_delay, arr_delay, method = "kendall")
```

If **not normally distributed** and there is a **large number of ties**, use **Kendall’s tau**

- non-parametric
- based on rank difference

The square of the correlation value indicates the percentage of shared variance

**Departure and arrival delay seem actually to share**

- `r flights_nov_20_cor_kendall$estimate %>% round(digits = 3)` ^ 2 = `r flights_nov_20_cor_kendall$estimate ^ 2 %>% round(digits = 3)`
- **`r ((flights_nov_20_cor_kendall$estimate ^ 2) * 100) %>% round(digits = 1)`% of variance**


```{r, echo=TRUE, message=FALSE}
flights_nov_20 %$%
  cor.test(
    dep_delay, arr_delay, 
    method = "kendall")
```

</div>



## Pairs plot

Combines in one visualisation: histograms, scatter plots, and correlation values for a set of variables

<div class="columns-2">

```{r, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}
library(psych)

flights_nov_20 %>%
  select(
    dep_delay, 
    arr_delay, 
    air_time
  ) %>%
  pairs.panels(
    method = "kendall"
  )
```

<center>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 4}
library(psych)

flights_nov_20 %>%
  select(dep_delay, arr_delay, air_time) %>%
  pairs.panels(method = "kendall")
```
</center>

</div>



## Summary

Correlation

- Pearson’s r
- Spearman’s rho
- Kendall’s tau
- Pairs plot

**Next**: Data transformations

- Z-scores
- Logarithmic transformations