```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
rm(list = ls())
```



# Hierarchical and density-based clustering



## Recap

**Prev**: Centroid-based clustering

- K-means
- Fuzzy c-means
- Geodemographic classification

**Now**: Hierarchical and density-based clustering

- Hierarchical
- Mixed 
- Density-based


## Libraries

```{r, echo=TRUE}
library(tidyverse)
library(e1071)
library(dbscan)
```


## Example

```{r, echo=TRUE, message=FALSE, warning=FALSE}
data_to_cluster <- data.frame(
  x_values = c(rnorm(40, 5, 1), rnorm(60, 10, 1), rnorm(20, 12, 3)),
  y_values = c(rnorm(40, 5, 1), rnorm(60, 5, 3), rnorm(20, 15, 1)),
  original_group = c(rep("A", 40), rep("B", 60), rep("C", 20)) )
```



## Hierarchical clustering

<font size="4">	
**Algorithm**: each object is initialised as, then repeat

- join the two most similar clusters based on a distance-based metric
- e.g., Ward's (1963) approach is based on variance

until only one single cluster is achieved
</font>

```{r, echo=TRUE, message=FALSE, warning=FALSE}
hclust_result <- data_to_cluster %>%
  select(x_values, y_values) %>%
  dist(method="euclidean") %>%
  hclust(method="ward.D2")

data_to_cluster <- data_to_cluster %>%
  add_column(hclust_cluster = cutree(hclust_result, k=3))
```

## Clustering tree

This approach generates a clustering tree (dendrogram), which can then be *"cut"* at the desired height

```{r, echo=TRUE, message=FALSE, warning=FALSE}
plot(hclust_result) + abline(h = 30, col = "red")
```

## Hierarchical clustering result

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
data_to_cluster %>%
  ggplot(aes(
    x = x_values, y = y_values, 
    colour = original_group)) +
  geom_point() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#e41a1c", "#377eb8", "#4daf4a")) +
  coord_fixed(ratio = 1)
```

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
data_to_cluster %>%
  ggplot(aes(
    x = x_values, y = y_values, 
    colour = factor(hclust_cluster))) +
  geom_point() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3")) +
  coord_fixed(ratio = 1)
```

:::
::::::


## Bagged clustering

Bootstrap aggregating (*b-agg-ed*) clustering approach (Leisch, 1999)

- first k-means on samples
- then a hierarchical clustering of the centroids generated through the samples

```{r, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}
bclust_result <- data_to_cluster %>%
  select(x_values, y_values) %>%
  bclust(hclust.method="ward.D2", resample = TRUE)

data_to_cluster <- data_to_cluster %>%
  add_column(bclust_cluster = clusters.bclust(bclust_result, 3))
```

## Bagged clustering result

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
data_to_cluster %>%
  ggplot(aes(
    x = x_values, y = y_values, 
    colour = original_group)) +
  geom_point() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#e41a1c", "#377eb8", "#4daf4a")) +
  coord_fixed(ratio = 1)
```

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
data_to_cluster %>%
  ggplot(aes(
    x = x_values, y = y_values, 
    colour = factor(bclust_cluster))) +
  geom_point() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3")) +
  coord_fixed(ratio = 1)
```

:::
::::::



## Density based clustering

DBSCAN (*"density-based spatial clustering of applications with noise"*) starts from an unclustered point and proceeds by aggregating its neighbours to the same cluster, as long as they are within a certain distance. (Ester *et al*, 1996)

```{r, echo=TRUE, message=FALSE, warning=FALSE}
dbscan_result <- data_to_cluster %>%
  select(x_values, y_values) %>%
   dbscan(eps = 1, minPts = 5)

data_to_cluster <- data_to_cluster %>%
  add_column(dbscan_cluster = dbscan_result$cluster)
```

## DBSCAN result

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
data_to_cluster %>%
  ggplot(aes(
    x = x_values, y = y_values, 
    colour = original_group)) +
  geom_point() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#e41a1c", "#377eb8", "#4daf4a")) +
  coord_fixed(ratio = 1)
```

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
data_to_cluster %>%
  ggplot(aes(
    x = x_values, y = y_values, 
    colour = factor(dbscan_cluster))) +
  geom_point() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#666666", "#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#e6ab02", "#a6761d")) +
  coord_fixed(ratio = 1)
```

:::
::::::



## Summary

Hierarchical and density-based clustering

- Hierarchical
- Mixed 
- Density-based

**Next**: Practical session

- Geodemographic classification