```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("GRANOLARR_HOME"))
rm(list = ls())
```



# Data manipulation



## Recap

**Prev**: Data selection and filtering

- dplyr
- dplyr::select
- dplyr::filter

**Now**: Data manipulation

- dplyr::arrange
- dplyr::summarise
- dplyr::group_by
- dplyr::mutate



## dplyr

The `dplyr` (pronounced *dee-ply-er*) library is part of `tidyverse` and it offers a grammar for data manipulation

- `select`: select specific columns
- `filter`: select specific rows
- `arrange`: arrange rows in a particular order
- `summarise`: calculate aggregated values (e.g., mean, max, etc)
- `group_by`: group data based on common column values
- `mutate`: add columns
- `join`: merge data frames


## Libraries

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(nycflights13)

nov_dep_delays <- 
  nycflights13::flights %>%
  select(origin, dest, dep_delay, arr_delay, year:day) %>%
  select(-arr_delay) %>%
  filter(month == 11)
```



## dplyr::arrange

```{r, echo=TRUE}
nov_dep_delays <- nov_dep_delays %>%
  arrange( 
    dest, # Ascending destination name
    -dep_delay # Descending delay
  )

nov_dep_delays %>% head(3)
```



## dplyr::summarise

`summarise`: calculate aggregated values (e.g., mean, max, etc)

```{r, echo=TRUE}
aggr_dep_delays_nov <- nov_dep_delays %>%
  # Need to filter out rows where delay is NA
  filter(!is.na(dep_delay)) %>%
  # Create two aggregated columns
  summarize(
    avg_dep_delay = mean(dep_delay), 
    tot_dep_delay = sum(dep_delay)
  )

aggr_dep_delays_nov
```



## dplyr::group_by

```{r, echo=TRUE}
dest_dep_delays_nov <- nov_dep_delays %>%
  # Need to filter out rows where delay is NA
  filter(!is.na(dep_delay)) %>%
  # Frist group by same destination
  group_by(dest) %>%
  # Then caluculate aggregated value
  summarize(
    tot_dep_delay = sum(dep_delay)
  )

dest_dep_delays_nov %>% head(3)
```



## dplyr::mutate

```{r, echo=TRUE}
dest_dep_delays_nov <- dest_dep_delays_nov %>%
  mutate(
    tot_dep_delay_days = ((tot_dep_delay / 60) /24)
  )

dest_dep_delays_nov %>% head(3)
```


## Full pipe example

```{r, echo=TRUE}
dest_dep_delays_nov <- nycflights13::flights %>%
  select(origin, dest, dep_delay, arr_delay, year:day) %>%
  select(-arr_delay) %>%
  filter(month == 11) %>%
  filter(!is.na(dep_delay)) %>%
  arrange(dest, -dep_delay) %>%
  group_by(dest) %>%
  summarize(tot_dep_delay = sum(dep_delay)) %>%
  mutate(tot_dep_delay_days = ((tot_dep_delay / 60) /24))
  
dest_dep_delays_nov %>% head(3)
```



## Summary

Data manipulation

- dplyr::arrange
- dplyr::summarise
- dplyr::group_by
- dplyr::mutate

**Next**: Practical session

- Creating R projects
- Creating R scripts
- Data wrangling script

```{r cleanup, include=FALSE}
rm(list = ls())
```