```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("GRANOLARR_HOME"))
rm(list = ls())
```



# Read and write data



## Summary

Table pivot

- Wide and long data
- tidyr::pivot_longer
- tidyr::pivot_wider

**Next**: Read and write data

- file formats
- read
- write



## Comma Separated Values

The file `2011_OAC_Raw_uVariables_Leicester.csv`
- contains data used for the 2011 Output Area Classificagtion
- 167 variables, as well as the resulting groups
- for the city of Leicester

Extract showing only some columns

```{}
OA11CD,LSOA11CD, ... supgrpcode,supgrpname,Total_Population, ...
E00069517,E01013785, ... 6,Suburbanites,313, ...
E00169516,E01013713, ... 4,Multicultural Metropolitans,341, ...
E00169048,E01032862, ... 4,Multicultural Metropolitans,345, ...
```

The full variable names can be found in the file
- `2011_OAC_Raw_uVariables_Lookup.csv`.


## Libraries

```{r, echo=TRUE}
library(tidyverse)
library(knitr)
```


## Read

The `read_csv` function of the [`readr`](https://readr.tidyverse.org/index.html) library reads a *csv* file from the path provided as the first argument


```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC <- read_csv("2011_OAC_Raw_uVariables_Leicester.csv")

leicester_2011OAC %>% 
  select(OA11CD,LSOA11CD, supgrpcode,supgrpname,Total_Population) %>%
  top_n(3) %>%
  kable()
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
leicester_2011OAC <- read_csv(paste0(Sys.getenv("GRANOLARR_HOME"), "/data/", "2011_OAC_Raw_uVariables_Leicester.csv"))
leicester_2011OAC %>% 
  select(OA11CD,LSOA11CD, supgrpcode,supgrpname,Total_Population) %>%
  top_n(3) %>%
  kable()
```


## Write

The function `write_csv` can be used to save a dataset to `csv`

Example:

1. **read** the 2011 OAC dataset
2. **select** a few columns
3. **filter** only those OA in the supergroup *Suburbanites* (code `6`)
4. **write** the results to a file named *Leicester_Suburbanites.csv* in your home folder

```{r, echo=TRUE, eval=FALSE}
read_csv("2011_OAC_Raw_uVariables_Leicester.csv") %>%
  select(OA11CD, supgrpcode, supgrpname) %>%
  filter(supgrpcode == 6) %>%
  write_csv("~/Leicester_Suburbanites.csv")
```


## Summary

Read and write data

- file formats
- read
- write

**Next**: Practical session

- Join operations
- Table pivot
- Read and write data

```{r cleanup, include=FALSE}
rm(list = ls())
```