```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("GRANOLARR_HOME"))
rm(list = ls())
```


<style type="text/css">
.small_r_all pre{
  font-size: 16px;
  line-height: 18px;
}
.small_r_output pre:not(.prettyprint){
  font-size: 16px;
  line-height: 18px;
}
.verysmall_r_output pre:not(.prettyprint){
  font-size: 12px;
  line-height: 14px;
}
</style>



# Comparing groups


## Recap

**Prev**: Exploratory data analysis

- 301 Lecture Data visualisation
- 302 Lecture Descriptive statistics
- 303 Lecture Exploring assumptions
- 304 Practical session

**Now**: Comparing groups

- T-test
- ANOVA
- Chi-square


## Libraries

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

Today's libraries

- mostly working with the usual `nycflights13`
- exposition pipe `%$%` from the library `magrittr`

```{r, echo=TRUE, message=FALSE, warning=FALSE,}
library(tidyverse)
library(magrittr)  
library(nycflights13)
```

:::

::: {.col data-latex="{0.5\textwidth}"}

But let's start from a simple example from `datasets`

- 50 flowers from each of 3 species of iris

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3.5, fig.height = 5}
iris %>%
  ggplot(
    aes(
      x = Species, 
      y = Petal.Length
    )
  ) +
  geom_boxplot()
```

:::
::::::



## Example

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
iris %>% filter(Species == "setosa") %>% pull(Petal.Length) %>% shapiro.test()

iris %>% filter(Species == "versicolor") %>% pull(Petal.Length) %>% shapiro.test()

iris %>% filter(Species == "virginica") %>% pull(Petal.Length) %>% shapiro.test()
```

</div>



## T-test

Independent T-test tests whether two group means are different

$$outcome_i = (group\ mean) + error_i $$

- groups defined by a predictor, categorical variable
- outcome is a continuous variable
- assuming
    - normally distributed values in groups
    - homogeneity of variance of values in groups
        - if groups have different sizes
    - independence of groups



## Example

Values are normally distributed, groups have same size, and they are independent (different flowers, check using `leveneTest`)

<div class="small_r_output">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
iris %>%
  filter(Species %in% c("versicolor", "virginica")) %$% # Note %$%
  t.test(Petal.Length ~ Species)
```

</div>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
iris_t_test <- iris %>%
  filter(Species %in% c("versicolor", "virginica")) %$%
  t.test(Petal.Length ~ Species)
```

The difference is significant t(`r iris_t_test %$% parameter[["df"]] %>% round(digits = 2)`) = `r iris_t_test %$% statistic[["t"]] %>% round(digits = 2)`, *p* < .01



## ANOVA

ANOVA (analysis of variance) tests whether more than two group means are different

$$outcome_i = (group\ mean) + error_i $$

- groups defined by a predictor, categorical variable
- outcome is a continuous variable
- assuming
    - normally distributed values in groups
        - especially if groups have different sizes
    - homogeneity of variance of values in groups
        - if groups have different sizes
    - independence of groups



## Example

Values are normally distributed, groups have same size, they are independent (different flowers, check using `leveneTest`)

<div class="small_r_output">

<!--
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(car)

iris %$%
leveneTest(Petal.Length, Species)
```
-->

```{r, echo=TRUE, message=FALSE, warning=FALSE}
iris %$%
  aov(Petal.Length ~ Species) %>%
  summary()
```

</div>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
iris_anova_summary <- iris %$%
  aov(Petal.Length ~ Species) %>%
  summary()
```

The difference is significant t(`r iris_anova_summary[[1]] %$% Df[1] %>% round(digits = 2)`, `r iris_anova_summary[[1]] %$% Df[2] %>% round(digits = 2)`) = `r iris_anova_summary[[1]][1, 4] %>% round(digits = 2)`, *p* < .01


## Summary

Comparing groups

- T-test
- ANOVA
- Chi-square

**Next**: Correlation

- Pearson’s r
- Spearman’s rho
- Kendall’s tau
- Pairs plot

```{r cleanup, include=FALSE}
rm(list = ls())
```