```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("GRANOLARR_HOME"))
rm(list = ls())
```

<style type="text/css">
.small_r_all pre{
  font-size: 16px !important;
  line-height: 18px !important;
}
.small_r_output pre:not(.prettyprint){
  font-size: 16px;
  line-height: 18px;
}
.verysmall_r_output pre:not(.prettyprint){
  font-size: 12px;
  line-height: 14px;
}
</style>



# Multiple Regression


## Recap

**Prev**: Simple Regression

- Regression
- Ordinary Least Squares
- Interpretation
- Checking assumptions

**Now**: Multiple Regression

- Multiple regression
- Ordinary Least Squares
- Interpretation
- Checking assumptions



```{r, echo=FALSE, message=FALSE, warning=FALSE,}
library(tidyverse)
library(magrittr)  
library(palmerpenguins)
library(MASS)
library(lmtest)
```


## Multiple regression

**Regression analysis** is a supervised machine learning approach

Special case of the general linear model

$$outcome_i = (model) + error_i $$

Predict (estimate) value of one outcome (dependent) variable as

- one predictor (independent) variable: **simple / univariate**

$$Y_i = (b_0 + b_1 * X_{i1}) + \epsilon_i $$
    
- more predictor (independent) variables: **multiple / multivar.**

$$Y_i = (b_0 + b_1 * X_{i1} + b_2 * X_{i2} + \dots + b_M * X_{iM}) + \epsilon_i $$




## Assumptions

- **Linearity**
    - the relationship is actually linear
- **Normality** of residuals
    - standard residuals are normally distributed with mean `0`
- **Homoscedasticity** of residuals
    - at each level of the predictor variable(s) the variance of the standard residuals should be the same (*homo-scedasticity*) rather than different (*hetero-scedasticity*) 
- **Independence** of residuals
    - adjacent standard residuals are not correlated
- When more than one predictor: **no multicollinearity**
    - if two or more predictor variables are used in the model, each pair of variables not correlated



## Boston housing

:::::: {.cols data-latex=""}

::: {.col style="width: 80%;" data-latex="{0.5\textwidth}"}

A classic R dataset

- price of (506) houses in Boston
- in relation to: 
  - house characteristics
  - neighborhood 
  - air quality

<br/>
Harrison, D., and D. L. Rubinfeld. 1978. [Hedonic Housing Prices and the Demand for Clean Air](https://doi.org/10.1016/0095-0696(78)90006-2). Journal of Environmental Economics and Management 5 (1): 81–102.

:::

::: {.col style="width: 70%; text-align: right;" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3.5, fig.height = 5}
MASS::Boston %>%
  ggplot2::ggplot(
    aes(
      x = nox, 
      y = medv
    )
  ) +
  xlab("Nitrogen oxides (NO) concentration") +
  ylab("Median value in $1,000") +
  ggplot2::geom_point()
```

:::
::::::




## Example

Can we predict price based on number of rooms and air quality?

$$house\ value_i = (b_0 + b_1 * rooms_{i} + b_1 * NO\ conc_{i}) + \epsilon_i $$

<center>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 4}
MASS::Boston %>%
  dplyr::select(medv, nox, rm) %>%
  psych::pairs.panels(
    method = "kendall",
    stars = TRUE
  )
```
</center>


## stats::lm

<div class="small_r_all">

```{r, echo=TRUE}
MASS::Boston %$% 
  stats::lm(medv ~ nox + rm) ->
  fit_medv_model

fit_medv_model %>%  
  summary()
```

</div>



## Overall fit

```{r, echo=FALSE}
fit_medv_model_summary <- fit_medv_model %>%
  summary()
```

The output indicates

- **p-value: < 2.2e-16**: $p<.01$ the model is significant
  - derived by comparing **F-statistic** to F distribution `r fit_medv_model_summary$fstatistic[1] %>% round(digits = 2)` having specified degrees of freedom (`r fit_medv_model_summary$fstatistic[2]`, `r fit_medv_model_summary$fstatistic[3]`)
  - Report as: F(`r fit_medv_model_summary$fstatistic[2]`, `r fit_medv_model_summary$fstatistic[3]`) = `r fit_medv_model_summary$fstatistic[1] %>% round(digits = 2)`
- **Adjusted R-squared: `r fit_medv_model_summary$adj.r.squared %>% round(digits = 4)`**: 
  - number of rooms and air quality can account for `r (fit_medv_model_summary$adj.r.squared * 100) %>% round(digits = 2)`% variation in house prices
- **Coefficients**
  - Intercept estimate `r fit_medv_model_summary$coefficients[1,1] %>% round(digits = 4)` is significant
  - `nox` (slope) estimate `r fit_medv_model_summary$coefficients[2,1] %>% round(digits = 4)` is significant
  - `rm` (slope) estimate `r fit_medv_model_summary$coefficients[3,1] %>% round(digits = 4)` is significant



## Outliers and influential cases

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
MASS::Boston %>%
  mutate(
    model_stdres = fit_medv_model %>% stats::rstandard(),
    model_cook_dist = fit_medv_model %>% stats::cooks.distance()
  ) ->
  boston_output

boston_output %>%
  dplyr::select(medv, model_stdres, model_cook_dist) %>%
  dplyr::filter(abs(model_stdres) > 2.58 | model_cook_dist > 1)
```

</div>

No influential cases (Cook's distance `> 1`) but there are many outliers (7 abs std res `> 3.29`, 2% `> 2.58`)



## Checking assumptions: normality

Shapiro-Wilk test for normality of standard residuals, 

- robust models: should be **not** significant 

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=TRUE, message=FALSE, warning=FALSE}
boston_output %$% 
  stats::shapiro.test(
    model_stdres
  )
```

<font size="4">	
**Standard residuals are NOT normally distributed**
</font>

:::

::: {.col data-latex="{0.5\textwidth}"}

<center>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 3}
boston_output %>%
  ggplot2::ggplot(aes(x = model_stdres)) +
  ggplot2::geom_histogram(
    aes(
      y =..density..
    ),
    bins = 100
  ) + 
  ggplot2::stat_function(
    fun = dnorm, 
    args = list(
      mean = boston_output %>% pull(model_stdres) %>% mean(),
      sd = boston_output %>% pull(model_stdres) %>% sd()),
    colour = "red", size = 1)
```
</center>

:::
::::::

</div>



## Checking assumpt.: homoscedasticity

Breusch-Pagan test for homoscedasticity of standard residuals

- robust models: should be **not** significant

<div class="small_r_output">

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 3}
fit_medv_model %>% 
  lmtest::bptest()
```

<font size="4">	
**Standard residuals are NOT homoscedastic**
</font>

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 3}
fit_medv_model %>% 
  plot(which = c(1))
```

:::
::::::

</div>



## Checking assumptions: independence

Durbin-Watson test for the independence of residuals

- robust models: statistic should be close to 2 (advised between 1 and 3) and **not** significant

<div class="small_r_output">

```{r, echo=TRUE}
fit_medv_model %>%
  lmtest::dwtest()
```

</div>

<font size="4">	
**Standard residuals are NOT independent**

Note: the result depends on the order of the data.
</font>



## Checking assumpt.: multicollinearity

Checking the variance inflation factor (VIF)

- robust models should have no multicollinearity: 
  - largest VIF should be lower than 10 or the average VIF should not be greater than 1


```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(car)

fit_medv_model %>%
  car::vif()
```

<font size="4">	
**There is no multicollinearity**
</font>



## Example

No, we can't predict house prices based only on number of rooms and air quality.

- predictors are statistically significant
- but model is not robust, as it doesn't satisfy most assumptions
  - Standard residuals are NOT normally distributed
  - Standard residuals are NOT homoscedastic
  - Standard residuals are NOT independent
  - (although there is no multicollinearity)

We seem to be on the right path, but something is missing...


## Example

<div class="small_r_all">

```{r, echo=TRUE}
MASS::Boston %$% 
  stats::lm(medv ~ nox + rm + ptratio + crim) ->
  fit_medv_model2

fit_medv_model2 %>%  
  summary()
```

</div>




## Information criteria

<!--
$R^2$ values tend to increase with the number of variables included in the model

- Adjusted $R^2$ 
  - controls for number of cases and variables
  - interpretable
-->

- Akaike Information Criterion (**AIC**)
  - measure of model fit 
    - penalising model with more variables
  - not interpretable per-se, used to compare models
    - lower value, better fit
- Bayesian Information Criterion (**BIC**)
  - similar to AIC
  
<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
stats::AIC(fit_medv_model)
stats::AIC(fit_medv_model2)
```

</div>


## Checking assumptions

<div class="small_r_all">

```{r, echo=FALSE, message=FALSE, warning=FALSE}
fit_medv_model2 %>% 
  stats::rstandard() %>% 
  stats::shapiro.test()

fit_medv_model2 %>% 
  lmtest::bptest()

fit_medv_model2 %>%
  lmtest::dwtest()

fit_medv_model2 %>%
  car::vif()
```

</div>


## Systematic choice

**Stepwise selection** of predictor (independent) variables

- iteratively adding and/or removing predictors 
- to obtain best performing model

Three approaches

- forward selection: from no variable, iteratively add variables
- backward selection: from all variables, iteratively remove variables
- step-wise (both) selection: 
  - from a given model
  - one step forward, add most promising variable
  - one step backward, remove any variable not improving


<!--
## Outliers and residuals
## Influential cases
-->



## Summary

Simple Regression

- Regression
- Ordinary Least Squares
- Fit

**Next**: Assessing regression assumptions

- Normality
- Homoscedasticity
- Independence

```{r cleanup, include=FALSE}
rm(list = ls())
```