```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("GRANOLARR_HOME"))
rm(list = ls())
```

<style type="text/css">
.small_r_all pre{
  font-size: 16px;
  line-height: 18px;
}
.small_r_output pre:not(.prettyprint){
  font-size: 16px;
  line-height: 18px;
}
.verysmall_r_output pre:not(.prettyprint){
  font-size: 12px;
  line-height: 14px;
}
</style>


# Assessing regression assumptions


## Recap

**Prev**: Simple Regression

- Regression
- Ordinary Least Squares
- Fit

**Now**: Assessing regression assumptions

- Normality
- Homoscedasticity
- Independence


## Checking assumptions

- **Linearity**
    - the relationship is actually linear
- **Normality** of residuals
    - standard residuals are normally distributed with mean `0`
- **Homoscedasticity** of residuals
    - at each level of the predictor variable(s) the variance of the standard residuals should be the same (*homo-scedasticity*) rather than different (*hetero-scedasticity*) 
- **Independence** of residuals
    - adjacent standard residuals are not correlated
- When more than one predictor: **no multicollinearity**
    - if two or more predictor variables are used in the model, each pair of variables not correlated


## Libraries and data

```{r, echo=TRUE, message=FALSE, warning=FALSE,}
library(tidyverse)
library(magrittr)  
library(nycflights13)

flights_nov_20 <- nycflights13::flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay), month == 11, day ==20) 
```


## Example

<font size="4">	
$$arr\_delay_i = (b_0 + b_1 * dep\_delay_{i1}) + \epsilon_i $$
</font>

<div class="small_r_output">

```{r, echo=TRUE}
delay_model <- flights_nov_20 %$% # Note %$%
  lm(arr_delay ~ dep_delay)

delay_model %>%  summary()
```

</div>




## Normality

Shapiro-Wilk test for normality of standard residuals, 

- robust models: should be not significant 

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=TRUE, message=FALSE, warning=FALSE}
delay_model %>% 
  rstandard() %>% 
  shapiro.test()
```

<font size="4">	
**Standard residuals are NOT normally distributed**
</font>

:::

::: {.col data-latex="{0.5\textwidth}"}

<center>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 4}
delay_model %>% 
  rstandard() %>%
  data.frame(std_res = .) %>%
  ggplot(aes(x = std_res)) +
  geom_histogram(
    aes(
      y =..density..
    ),
    bins = 100
  ) + 
  stat_function(
    fun = dnorm, 
    args = list(
      mean = delay_model %>% rstandard() %>% mean(),
      sd = delay_model %>% rstandard() %>% sd()),
    colour = "red", size = 1)
```
</center>

:::
::::::

</div>



## Homoscedasticity

Breusch-Pagan test for homoscedasticity of standard residuals

- robust models: should be not significant

<div class="small_r_output">

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 3}
library(lmtest)

delay_model %>% 
  bptest()
```

<font size="4">	
**Standard residuals are homoscedastic**
</font>

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 3}
delay_model %>% 
  plot(which = c(1))
```

:::
::::::

</div>



## Independence

Durbin-Watson test for the independence of residuals

- robust models: statistic should be close to 2 (between 1 and 3) and not significant

<div class="small_r_output">

```{r, echo=TRUE}
# Also part of the library lmtest
delay_model %>%
  dwtest()
```

</div>

<font size="4">	
**Standard residuals might not be completely indipendent**

Note: the result depends on the order of the data.
</font>




## Summary

Assessing regression assumptions

- Normality
- Homoscedasticity
- Independence

**Next**: Assessing regression assumptions

- Normality
- Homoscedasticity
- Independence

```{r cleanup, include=FALSE}
rm(list = ls())
```